\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-26 chapters/bci.tex
Add ERP section

\todo{comments liesel paradigms}
People suffering from heavy paralysis or disability due to
neurodegenenerative diseases, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

@@

The key problem for heavily paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
If the mind's seat is in the brain, and we want to interface this to a digital
control process, such a device becomes the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
assistive technology and communication devices.}.
A \ac{bci} directly couples actions in the `real world' to the mental state of
the user.
A \ac{bci} reads your mind.

Starting in the 1970's from methods to establish minimal brain activity
(yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing part of the nervous system and muscle control.
By now, classic \acp{bci} methods have matured outside the lab setting and can be
used as assistive communication technology by patients with impaired
speech~\cite{Wolpaw2018}.
Impressive cutting edge examples of experimental \acp{bci} include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
Neuralink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation and medical diagnosis or treatment, but
have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

Simply put, a \ac{bci} records the user's brain activity, then extracts some
relevant output from this brain activity and couples this output to a function
of a device.
Optionally, the user can then observe the action of the device and adapt their
behavior or brain activity accordingly, closing the loop
\footnote{%
  \textcite{BCISociety2024} has recently formalized this into the
  following definition:
  \it``A brain-computer interface is a system that measures brain activity and
  converts it in (nearly) real-time into functionally useful outputs to replace,
  restore, enhance, supplement, and/or improve the natural outputs of the brain,
  thereby changing the ongoing interactions between the brain and its external or
  internal environments. It may additionally modify brain activity using targeted
  delivery of stimuli to create functionally useful inputs to the
  brain.''
}.

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to specific brain
regions that are more or less active at a given time with a scanner.
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal \emph{action potential}.
The brain consists 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The action potential is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.\todo{gross oversimplification,not
entirely true}
A neuron's action potential, or the combined action potentials of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical potential can be faster and more
informative.

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the letter shown on the
stimulus.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signals related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own electrical signals.
On the other hand, the activity of interest in often
originates from a specific functional system within the brain.
This activity might not easily be discerned from all other neural and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
relevant signal is often solved through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\label{sec:bci-decoding} takes a closer look at some common decoding
techniques.

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offer great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use active electrodes with preamplifiers and
electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality, since they require a trained
professional to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

Reactive \ac{bci} take another approach, by giving the user a discrete set of
sensory stimulations and decoding towards which stimulus the user's attention
was directed.
Compared to active \ac{bci}, reactive paradigms usually require higher effort
and can induce fatigue due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g. speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

The major visual paradigms comprise
\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} type interfaces.
In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
among many that each flicker continuously at different
frequencies~\cite{Chen2021}.
In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
be correlated with the brain activity to retrieve the attended stimulus.
Instead of stimulating all possible selections at once, the oddball paradigm
stimulates them one by one with single flashing.\todo{reference}
Since the times of stimulation of each target is known by the interface,
information is modulated by time of stimulation, and times where an attention
response can be found correspond to selected targets.
\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
movements in specific directions instead of flashing.
Information is then also carried by movement
direction~\cite{Libert2021a,Libert2022}.

@@

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Some valuable metrics to asses decoder performance in the classification case
are accuracy and \ac{rocauc}.
Accuracy is calculated as the proportion of correct selections to all
selections made.
It should be carefully interpreted in the presence of imbalanced data and always compared to the
random chance accuracy level in the presence of more than two possible
selections per trial.
\Ac{rocauc} is also a measure of classifier predictive power, but is more
suited for evaluation of classification of single epochs of data and in the
presence of imbalanced data, e.g. when comparing single \acp{erp}.
Higher \ac{rocauc} usually translates to a higher target selection accuracy.

@@

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
Intensifcations of different targets are usually 100-200ms apart and last for about
100ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it flashes.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

@@

\subsection{The event-related potential}

>>>>

\todo{comments liesel paradigms}
People suffering from severe paralysis or disability due to
neurodegenenerative disease, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

@@

The key problem for heavily paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton. The term \ac{bci} is usually preferred for
assistive technology and communication devices.}.
A \ac{bci} directly couples actions in the `real world' to the mental state of
the user.
A \ac{bci} reads your mind.

Starting in the 1970's from methods to establish basic communication from
minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods have matured outside the lab setting and can be
used as assistive communication technology by patients with impaired
speech~\cite{Wolpaw2018}.
Impressive cutting edge examples of experimental \acp{bci} include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
Neuralink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

\section{A direct interface to the brain}

Brain-computer interfacing is an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation and medical diagnosis or treatment, but
have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

Simply put, a \ac{bci} records the user's brain activity, extracts some
relevant output from this brain activity and couples this output to a function
of a device.
Optionally, the user can then observe the action of the device and adapt their
behavior or brain activity accordingly, closing the loop
\footnote{%
  \textcite{BCISociety2024} has recently formalized this into the
  following definition:
  \it``A brain-computer interface is a system that measures brain activity and
  converts it in (nearly) real-time into functionally useful outputs to replace,
  restore, enhance, supplement, and/or improve the natural outputs of the brain,
  thereby changing the ongoing interactions between the brain and its external or
  internal environments. It may additionally modify brain activity using targeted
  delivery of stimuli to create functionally useful inputs to the
  brain.''
}.

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to establish the
activity of a specific brain region
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal \emph{action potential}.
The brain consists 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The action potential is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.\todo{gross oversimplification,not
entirely true}
A neuron's action potential, or the combined action potentials of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical potential can be faster and more
informative.

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the inteded letter A or B
in Figure~\cite{fig:bci/loop}.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest in often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\label{sec:bci-decoding} takes a closer look at some common decoding
techniques.

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offer great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use active electrodes with preamplifiers and
electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality, since they require a trained
individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

Reactive \ac{bci} take another approach, by providing the user a discrete set of
sensory stimulations and decoding towards which stimulus the user's attention
was directed.
Compared to active \ac{bci}, reactive paradigms usually require higher effort
and can induce fatigue due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g. speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

The major visual paradigms rely on
\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} brain responses.
In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
among many that each flicker continuously at different
frequencies~\cite{Chen2021}.
In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
be correlated with the brain activity to retrieve the attended stimulus.
Instead of stimulating all possible selections at once, the oddball paradigm
stimulates them one by one with single flashing.\todo{reference}
Since the times of stimulation of each target is known, and time-points at
which an attention signature is detected to selected targets.
\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
movements in specific directions instead of flashing.
Information is then also carried by movement
direction~\cite{Libert2021a,Libert2022}.

@@

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Accuracy is a valuable metrics to asses decoder performance in the classification
case.
Accuracy is calculated as the proportion of correct selections to all
selections made.
It should be carefully interpreted in the presence of imbalanced data and always compared to the
random chance accuracy level in the presence of more than two possible
selections per trial.
Alternatively, \ac{rocauc} is also a measure of classifier predictive power, but is more
suited for evaluation of classification of single epochs of data and in the
presence of imbalanced data, e.g. when comparing single \acp{erp}.
Higher \ac{rocauc} usually translates to a higher target selection accuracy.

@@

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
Intensifcations of different targets are usually 100-200ms apart and last for about
100ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it occurs.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

@@

\subsection{The \acf{erp}}
